

```{r build, eval=FALSE, include=FALSE}
knitr:::opts_chunk$set(collapse=TRUE)
knitr:::opts_chunk$set(tidy.opts=list(width.cutoff=60))
knit2=function(file){
    source("frame.R")
    require(knitr); require(rmarkdown)
    fse=tools:::file_path_sans_ext(file)
    knit(file)
    render(paste0(fse, '.md'), "html_document")
    render(paste0(fse, '.md'), pdf_document())
}
knit2("README.Rmd")
 

## with with markdown lib
## markdownToHTML("README.md", "README.html")
## shell("pandoc README.md -o README.pdf", shell=Sys.getenv("COMSPEC"))

```


'frame' an S4 class inheriting from data.frame
==============================================

Find me on [github](https://github.com/AntonioFasano/frame)

A class inheriting from data.frame featuring  `drop=FALSE` as default, an 'end' keyword, 0 index for 'all' and a 'desc' field.


The problem
-----------

Despite very powerful, R language is mostly intended for interactive use. So in `x[i,j]`  it is not straight  to set  such a value for `i` or `j` to emulate  `x[,j]` or `x[i,]`.

Using `TRUE` (which will get recycled)  is a trick to obtain this. 



```{r trick.matrix}

T=TRUE

x=matrix(1:12, ncol=3)
identical(x[,3], x[T,3])
identical(x[,3], `[`(x,T,3))

z=y=x
x[,3] =0
y[T,3]=0
identical(x,y)
z=`[<-`(z,T,3,0)
identical(x,z)

```
    
This works also for a data.frame. 

```{r trick.df}

x=data.frame(matrix(1:12, ncol=3))
identical(x[,3], x[T,3])
identical(x[,3], `[`(x,T,3))

z=y=x
x[,3] =0
y[T,3]=0
identical(x,y)
z=`[<-`(z,T,3,0)
identical(x,z)

```

Also the default `drop=TRUE` feature means breaking some programmatic code, like simple `nrow`, `ncol`, which will fail if the dimensions are dropped.

All these features are blessed time savers in interactive code, but oblige to many if/else check, prone to errors, in programmatic use.


The solution
-----------


I developed  a new data.frame S4 class, `frame`, inheriting  from data.frame but exploiting the `0`, unused in subsetting, to select all rows or columns, so `x[0,j]` and `x[i,0]`  as   `x[,j]` and `x[i,]`.
Also  `drop=FALSE`  is used as the default value in place of `drop=TRUE`.
Besides it uses a Matlab like `end` operator to get the last  rows or columns item, e.g.: `x[3:end, 2:end]`.
A `desc` field can be added to annotate any  data description.



The old data.frame and the new one
---------------------------------

This class inherits from the data.frame class: if a function  works  with a data.frame it is supposed to work with 'frame'. If your code requires a formal data.frame, you can obtain it from  `x` frame with:

    dfm(x)

This is a shortcut for `as.data.frame` which in turn will call the specialised function `as.data.frame.frame`.



What you get on the field
-------------------------

```{r create.df}

## Create a frame, from scratch (with column names)
frame(x, col.names=letters[1:3], desc="Hello frame")
frame(1:4, 5:8, 9:12, desc="Hello frame")

## or use a data.frame
x=data.frame(1:4, 5:8, 9:12)
frame(x, desc="Hello frame")




```
Note the option for frame only `col.names`. If you don't use it,   data.frame by default checks column names, frame does not, as it  gives a somewhat unpleasant view. If you need it,   use `frame(1:4, 5:8, 9:12, desc="Hello frame", check.names = TRUE)`. 
Similarly a very annoying feature of data.frame is the automatic conversion of strings to factors. Here instead:



```{r stringsAsFactors}

x=frame(a=1:26, b=letters)
class(x$b) 
## ... while
x=data.frame(a=1:26, b=letters)
class(x$b) # unless you modified general R options


```

You migh find convenient start from  matrices:

```{r create.mat}

frame(matrix(1:90, ncol=3), desc='Hello frame')
as.frame(matrix(1:90, ncol=3), desc='Hello frame')
## use "col.names" argument, if you don't like the automatic naming 
```

Note the `...` signalling that only the first 20 rows were printed. 
Nothing is more boring than printing an object just to discover you get a useless time consuming  gigantic print.  

Let us see the standard subsetting. 

```{r s-subsetting}
(fr=as.frame(matrix(1:12, ncol=3)))

## Standard subsetting
fr[c(1,3),2:3]
fr[2,]
fr[2,c('V1', 'V3')]
fr[c(T,F,T,T),T]
# fr[2:3]  # error not supported 
```


New features: subsetting without dropping dimensions by default!


```{r ns-subsetting}

## Non-standard behaviour:
## drop is false by default
fr[1,1]
fr[1,1, drop=T] #Traditional behaviour
fr[ ,1, drop=T] #Traditional behaviour

New features: `end` operator.
Inside brackets `end` is a keyword and it means the last row or column according to its position. 


```{r end-subsetting}
(fr=as.frame(matrix(1:12, ncol=3)))
fr[end,end]  # 'end' is not quoted inside brackets!
## fr[end]   # error not supported
## fr[2:end] # error not supported
fr[2:end,end:1]

## 'end' is always a keyword inside brackets
end=100
fr[, end]
```

Let us examine subsetting with zero index. As an index zero means all (like `TRUE`).


```{r zero-index}

## Zero means all
fr[0,1]
fr[1,0]
fr[0,0] # whole frame
```

Anyway sequences starting with zero are kept with the old meaning.  

```{r zero-index-seq}

## Traditional zero behaviour
## For a DF [0:3,0:2] is like [1:3,1:2], and for 'frame' too
fr[0:3,0:2] # Your old code won't break

```

Use in a programmatic environment
---------------------------------

Consider this code snippet

```{r size-snippet}
size=function(x)
         if(nrow(x)>100) print("I am a long series") else
                          print("I am a short series")

```

It will break when a data.frame loses a dimension, due to the default `drop=TRUE`.
Because `nrow` does not apply to atomic objects. 

```{r size-snippet2}

df=as.data.frame(matrix(1:12, ncol=3))
fr=frame(df)

## All columns
size(df)

## Take only the first col (we lose a dimension)
# size(df[,1])
# Error: argument is of length zero

## Take only the first col (now we don't lose original dimension)
size(fr[,1])

```	

Assume that `fr` represents real observations.

```{r covariance}
fr=as.frame(matrix(rnorm(16), ncol=4))
```
Calculate the covariance matrix of some or all the `fr` columns. Generate a random frame like `fr`, calculate the same covariance matrix and obtain the differences.

```{r covariance2}

dcov=function(fr,i,j){
        v1=fr[i,j]
        m=matrix(rnorm(nrow(fr) * ncol(fr)), ncol=ncol(fr))
        v2=as.frame(m)[i,j]
        var(v1)-var(v2)
}

```    
Get delta-cov for  all columns in `fr`:

```{r covariance3}
dcov(fr, 0, 0)
```

Only columns starting from second:

```{r covariance4}
dcov(fr, 0, '2:end') #Quoting necessary!!
```

Note: Without quoting, if you have a variable `end=3`, you are sending `2:3`.

All columns, but remove last row:

```{r covariance5}
dcov(fr, '-end', 0) #Quoting necessary!!
```

'end' quotation
---------------

*Inside brackets* you *do not quote* 'end' to use it as a keyword for last element.

```{r quotation}

fr=as.frame(matrix(1:16, nrow=4))
end=100
fr[2:end,] #'end' is the last row (4 not 100)
```

*Outside brackets*, that is  in a programmatic environment when using 'end' keyword indirectly in a variable assignment, *quote* end!

```{r quotation2}
end=1
a=end
b='end'

## Without quoting, variable value 'end' is 1
fr[1:2,a]
## By quoting, variable value 'end' is a keyword
fr[1:2,b]
```

If the assignment is an *expression* contains 'end' the same rule applies to get it as a keyword: *quote too*.

```{r quotation3}
a='3:end'
b=0
fr[a,b]
```

**In assignments**, without quoting 'end', it will get the current value of end, if any and valid, or raise an error. Inside brackets it will be passed unevaluated and after converted to a keyword.
Quoting in brackets means querying for the row/column named "end".


**Bracket alternatives**. These rules keep for bracket alternatives `` `[`(x,i,j)`` and  `` `[<-`(x,i,j)``.

```{r Bracket alternatives}

`[`(fr,end,end)
a='3:end'
`[`(fr,a,a)

```


What if my column is named 'end'?
--------------------------------

```{r col-named-end}
fr=as.frame(matrix(1:16, nrow=4))
names(fr)[3]='end'
fr[1,]

```

Inside brackets nothing special happens. As a keyword 'end' is unquoted, while as a name:

```{r col-named-end2}
fr[, c('V1','end')]

```

For assignments to a variable, outside brackets, quote twice.

```{r col-named-end3}

a="'end'"
b="c('V1', 'end')"

fr[1:2,a]
fr[1:2,b]

```
Obviously you can quote twice also with `sQuote`, `dQuote` or escaping inner quotation marks with "\\".


